
```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# (PART\*) Cluster 101 {-}

# What is a Cluster?

A computing cluster is a set of many computers networked together. Because there are many computers working together, the network is able to handle computationally expensive tasks, like genome assemblies or advanced algorithms. Imagine you're building a house. It would take a long time by yourself! It's much better to have many builders working together.

Now that we have a team of workers, the next challenge is task management. A home construction team will need a manager to help delegate tasks. Similarly, the computing cluster uses management software to prioritize tasks, delegate workers (resources), and check on progress. The Fred Hutch cluster uses a common management and scheduling tool called [Slurm](https://slurm.schedmd.com/overview.html).

```{r, echo=FALSE, fig.alt='Your computer has 1 node, 8 CPUs, and 8 gigabytes of memory. The Hutch cluster has nearly 600 nodes, 4000 CPUs, and 40 terabytes of memory.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g149d37dd4a1_0_18")
```

How is the cluster different from a laptop or desktop? First, on your laptop you most likely interact with it using an operating system like Windows or MacOS. The Fred Hutch cluster uses a Linux operating system. Second, because many people use the cluster for many tasks, there isn't a central screen and keyboard. You access the cluster remotely from your computer! We will talk more about how to connect to the cluster in a [following chapter](#terminal).

<div class = "dictionary">
**Computing cluster**  
  
A set of computers networked together to perform large tasks.
</div>

# Account Setup

You will need an account to log in to the cluster. This ensures that data stays protected and the computing resources are shared fairly.

## Check your HutchNet ID

Your [HutchNet ID](https://centernet.fredhutch.org/cn/u/center-it/help-desk.html) is the standard login you receive when you start working at the Hutch or are an official affiliate. You can use it to login to most resources at Fred Hutch (Desktop Computer, Employee Self Service, VPN, Webmail) and our Scientific Computing systems. 

For example:  

- my email is `jsmith3@fredhutch.org`.  
- my HutchNet ID is `jsmith3`.  

<div class = "notice">
If one of your collaborators requires access to the Fred Hutch network you can submit a [non-employee action form](https://centernet.fredhutch.org/cn/f/hr/lcex/non-employee-action-form.html). Non-employee is a generic administrative term for affiliates, students, contractors, etc.
</div>

## Connect to a PI Account {#pi-account}

To use the Scientific Computing cluster at Hutch, your HutchNet ID must be associated with a PI account. The Scientific Computing Team (SciComp) tries to set users up with a connection to a PI account before they need it, but this is not automatic!  To ensure that you have been set up to use the cluster, please fill out [this Account Setup Form](https://forms.gle/5ct8mQCeBD7LUt6S7) and we will ensure you are set up correctly!

Now, let's connect to the cluster using a terminal!

# Terminal Setup {#terminal}

The next step is getting familiar with your Terminal. This is your portal to the cluster.

## What is a terminal?

The Terminal is a [command line interface](https://www.codecademy.com/article/command-line-interface). In other words, the Terminal is a software application that allows you to issue commands directly to your laptop or desktop computer. The Terminal is very useful because it allows you to run commands that don't have a graphical user interface (GUI). It can also connect you to computer networks, such as the Fred Hutch cluster! The Terminal setup is different depending on your operating system. Jump to the [Windows](#windows), [MacOS](#mac), or [Linux](#linux) sections below.

<div class = "terminal">
"Terminal" used to be synonymous with "computer". With the creation of operating systems like Windows and MacOS, computers became much easier to use and exploded in popularity! Your colleagues are almost always referring to the command line application when they say "Terminal".
</div>

## Windows Setup {#windows}

<details><summary>Click to view steps</summary><p>

You will need to install a Terminal application called PuTTY to connect to the Fred Hutch Cluster.

1. You should then see PuTTY available in the Software Center. Click "Install" and go through the Setup Wizard.

    ```{r, echo=FALSE, fig.alt='The Software Center application lists available apps, including PuTTY.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g15643d101eb_4_15")
    ```

    ```{r, echo=FALSE, fig.alt='The PuTTY page inside Software Center. The Install button is highlighted.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g15643d101eb_4_20")
    ```

    <div class = "notice">
    You can also [install PuTTY manually](faq.html#manual-putty) if you don't see it in the Software Center.
    </div>

1. PuTTY should now be available in your applications. Click on PuTTY to open.

    ```{r, echo=FALSE, fig.alt='The PuTTY application now appears in the Windows app menu.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g15643d101eb_4_28")
    ```

1. You should now see the PuTTY Configuration menu.

    ```{r, echo=FALSE, fig.alt='The PuTTY configuration menu is open.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g15643d101eb_4_35")
    ```

</p></details>

## Mac Setup {#mac}

<details><summary>Click to view steps</summary><p>

Mac machines come with a Terminal installed. 

1. Go to Finder > Applications > Utilities > Terminal and double-click. 

    ```{r, echo=FALSE, fig.alt='Screenshot of the Utilities directory. The Terminal app is highlighted.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g149d37dd4a1_0_9")
    ```

1. Your Terminal should look like this:
    
    ```{r, echo=FALSE, fig.alt='Screenshot of the Mac Terminal.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g149d37dd4a1_0_2")
    ```

</p></details>
    
## Linux Setup {#linux}

<details><summary>Click to view steps</summary><p>

The commonly used Linux distribution, Ubuntu, already comes with a Terminal installed. 

1. Press ctrl + alt + T. Your open Terminal window should look like this:

[SCREENSHOT]

1. Update the Terminal and prepare it for connecting to the cluster by running:

```
sudo apt install openssh-client
```

Enter your password and enter `Y` when prompted.

</p></details>

# Logging In

Now that you have your Terminal application ready, you want to connect to the cluster. You will do this using a method called [SSH](https://www.ssh.com/academy/ssh/protocol), which stands for "Secure SHell".

## What is `SSH`?

SSH is a secure way to remotely connect to another computer or network of computers. In other words, SSH helps us protect your data and the data on the Fred Hutch cluster through authentication.

<div class = "dictionary">
**Hostname**  
  
The hostname is the name, or label, assigned to a computer in a network. We are connecting to hostname `rhino.fhcrc.org` or `rhino` for short.
</div>

Before moving on, you will need to connect to the Fred Hutch wifi network, a networked ethernet jack, or the [Fred Hutch VPN](https://centernet.fredhutch.org/cn/u/center-it/help-desk/vpn.html). This is the first layer of security. 

The next set of steps are specific to your operating system.  

## Windows Login 

<details><summary>Click to view steps</summary><p>

1. Go to the PuTTY Configuration menu. Under "Host Name" type **rhino** and click "Open".

    ```{r, echo=FALSE, fig.alt='The PuTTY Host Name "rhino" has been typed in.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g15643d101eb_4_41")
    ```

1. You will be prompted to login. Type in your HutchNetID (e.g., `jsmith3`).

    ```{r, echo=FALSE, fig.alt='The login prompt for rhino.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g15643d101eb_4_48")
    ```
   
1. Enter your password. No`*` or symbols will show up, so type it in carefully!
1. You are now logged in! There should be a login message, with your name at the bottom.

    ```{r, echo=FALSE, fig.alt='Screenshot of the PuTTY Terminal.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g15643d101eb_4_60")
    ```

Congratulations! You are now logged in to the Fred Hutch cluster!

</p></details>

## Mac Login

<details><summary>Click to view steps</summary><p>

1. Type the following commands, substituting in your HutchNet ID:  

    ```
    ssh -Y HutchID@rhino
    ```
1. You will see a message that looks like `The authenticity of host 'rhino (XXX.XXX.XX.XX)' can't be established.` Type in `yes` and hit enter.
1. Enter your password. No`*` or symbols will show up, so type it in carefully!
1. You are now logged in! There should be a login message, with your name at the bottom.

    ```{r, echo=FALSE, fig.alt='Screenshot of the Mac Terminal.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g149d37dd4a1_0_43")
    ```
    
Congratulations! You are now logged in to the Fred Hutch cluster!

</p></details>
    
## Linux Login

<details><summary>Click to view steps</summary><p>

Congratulations! You are now logged in to the Fred Hutch cluster!

</p></details>

# Submit Your First Job

The strength of a computing cluster is the ability to do many jobs in parallel or on computers with more computing power than you have on your local computer. Thus the best way to use the cluster is to create short snippets of directions (a script) that a computer can perform without human input and then tell the computers to do these jobs.  

Now that you've logged into rhino you will be able to send your script to the networked computers in the cluster, that are collectively called gizmo. The Fred Hutch cluster uses [Slurm](https://slurm.schedmd.com/overview.html) to organize and prioritize jobs sent to gizmo. Based on your instructions in your script, Slurm will find a computer to run your job along with all the other requests from other users.

## Download the Script

We can use the `wget` command to download a script from GitHub. This means we don't have to write the script from scratch. Copy and paste the following into the terminal, and hit return:

```
wget https://raw.githubusercontent.com/FredHutch/slurm-examples/master/introduction/1-hello-world/01.sh
```

```{r, echo=FALSE, fig.alt='Screenshot of wget command output, showing successful file download.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g1579ffd7b01_0_0")
```

## Confirm the Download

Let's confirm that we can see the file we just downloaded. We can use the `ls` (list files) command for this. Type `ls` and hit return. You should see the file `01.sh` in your home directory. The `.sh` ending means this is a script meant to run from the command line. 

```
ls
```

```{r, echo=FALSE, fig.alt='Screenshot of ls command showing that the file is in the working directory.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g1579ffd7b01_0_6")
```

## Inspect the Script

Let's next inspect the script. The `cat` command, followed by a file name, lists the entire contents of a specific file.

```
cat 01.sh
```

```{r, echo=FALSE, fig.alt='Screenshot of cat command showing the two lines contained in the script described below.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g1579ffd7b01_12_2")
```

- The first line of the script, `#!/bin/bash`, indicates that this is a command line or "bash" script. 
- The second line is empty, and the third line, `echo "Hello, World"` means that the computer will "echo", or print out, "Hello, World".

## Submit the Script

We use the `sbatch` command to submit a script and start running a job on the cluster. Copy the following and hit return. You should see a message like "`Submitted batch job 12345678`". Your number will vary because this is a unique job identifier.

```
sbatch 01.sh
```

## Check the Output

Type `ls` again. You should now see a log file like `slurm-12345678.out` listed alongside your script `01.sh`. Let's use `cat` to inspect the output in the log file. We should see our message has been printed!

```
cat slurm-<your-number-here>.out
```

```{r, echo=FALSE, fig.alt='Screenshot of ls and cat command shows that the Hello World message has successfully printed.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g1579ffd7b01_12_22")
```

<div class = "dictionary">
`ls` 
  
This command lists the files in the current directory.

`cat` _filename_

This command prints the contents of a specific file <filename>.

`sbatch` _filename.sh_

This command submits a job to the cluster with instructions specified in <filename>.sh
</div>

# File Upload and Download

Exchanging files with the cluster is very important. You can imagine scenarios where:

- You want to download log files or output files
- You want to upload a custom `.sh` script file that you wrote on your laptop
- You want to upload other files

In this course, upload and download of files is performed using [Cyberduck](https://cyberduck.io/). Cyberduck is a tool that lets us connect to the cluster securely, browse files, and transfer files securely.

## Download Cyberduck

Download the latest version of Cyberduck [here](https://cyberduck.io/download/).

<div class = "warning">
Note that the version of Cyberduck in the Software Center or Self Service might not be current, causing compatibility issues with some operating systems.
</div>

## Create Connection

Launch Cyberduck and click on "Open Connection".

- From the dropdown menu, select "SFTP (SSH File Transfer Protocol)"
- For Server, type "rhino.fhcrc.org"
- Fill in your HutchNetID for Username and fill in your password

Click "Connect"

```{r, echo=FALSE, fig.alt='Screenshot of Cyberduck “Open Connection” configuration.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g1579ffd7b01_12_28")
```

Click "Allow". You can also check the box to indicate "Always".

```{r, echo=FALSE, fig.alt='You can select allow when Cyberduck asks about an Unknown Fingerprint.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g1579ffd7b01_12_33")
```

You should see your script file "`01.sh`" and the log file.

```{r, echo=FALSE, fig.alt='Files, including the script we previously ran, can be accessed via Cyberduck.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g1579ffd7b01_12_37")
```

## Download and Edit the Script

- Right click on "`01.sh`" and select "Download"
- You will see a "Transfers" prompt open, and the `01.sh` file should now appear in your Downloads folder
- Open the `01.sh` file

```{r, echo=FALSE, fig.alt='Screenshot of the script opened on the users local laptop.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g1579ffd7b01_12_41")
```

Edit the message to include your name and save the file. Rename the file `01-name.sh`

```{r, echo=FALSE, fig.alt='Screenshot of the edited and renamed script file contents. The contents of the script now reads "Hello, Ava!".', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g1579ffd7b01_12_45")
```

## Upload the New Script

From your Downloads folder, simply drag the file to Cyberduck.

```{r, echo=FALSE, fig.alt='Screenshot of local and Cyberduck files, depicting the dragging that transfers the file over.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g1579ffd7b01_12_49")
```

You should now see the new script among your cluster files.

```{r, echo=FALSE, fig.alt='The new file is now listed in Cyberduck and is therefore on the cluster.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g15cf3fa00a4_0_6")
```

## Run the New Script

Return to your Terminal. Submit a job with your new script by running the following. When you type `ls` you should see a new log file! 

```
sbatch 01-name.sh
```

```{r, echo=FALSE, fig.alt='After running the newly edited script, there is a new log file present on the cluster.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.gff2211b72f_1_0")
```

<div class = "notice">
The job numbers included in log file names generally increase in number. The greater the number, the more recently the job was run.
</div>

Use the `cat` command to inspect the log. The message should show the new text that you added!

```
cat slurm-<your-number-here>.out
```

```{r, echo=FALSE, fig.alt='The message "Hello, Ava!" has been printed to the Terminal using the cat command.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.gff2211b72f_1_6")
```

# Interactive Session

While using the cluster, you might need to build and test jobs interactively before running them on gizmo. Luckily, you can launch an interactive session on the cluster by asking for part of a computer, called a "node", to be dedicated to you for a period of time, rather than using the Slurm job submission system. Interactive nodes are also useful if you have to install specific software for testing prior to requesting a software module be installed. 

## Starting the Session

When starting an interactive session, you're going to need to think about what you are testing and what resources you might need on the node you are requesting to use. We'll describe what these choices mean, though you can always start a session using the default values if you aren't sure what you need yet.  

Start an interactive node by running the command:

```
grabnode
```

You will be prompted with several questions about the type of interactive node you want. We don't need anything fancy, so we will set up the node to use minimal resources. You can enter the following:

- _How many CPUs/cores would you like to grab on the node?_ **1**
- _How much memory (GB) would you like to grab?_ **20**
- _Please enter the max number of days you would like to grab this node:_ **1**
- _Do you need a GPU?_ **N**
- When prompted, enter your password

The CPU, or Central Processing Unit, is the brain of the computer that performs and orchestrates computational tasks. Modern computers often perform multiple tasks at once, ranging from 4 tasks on a typical laptop to 48 tasks or more on higher end servers. 

RAM, or Random Access Memory, is often simply referred to as memory. This short term memory holds the information that the CPU needs to perform calculations. One distinctive feature of memory is that it is short term. In other words, when the electricity is shut off, the data stored in memory disappears. To save the CPU's work, you usually save files to your computer. Running highly complicated analyses or algorithms can often require additional memory resources.

The [GPU](https://www.intel.com/content/www/us/en/products/docs/processors/what-is-a-gpu.html), or Graphics Processing Unit, is similar to the CPU. The GPU was originally designed to quickly render graphics (such as for video games), but today can be used to run complex artificial intelligence applications or computationally intensive jobs.

```{r, echo=FALSE, fig.alt='CPU is the brain of the computer, whereas memory stores information in the short term.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.gff2211b72f_1_26")
```

You will see that you are now logged on to a "gizmo" node instead of a "rhino" node.

```{r, echo=FALSE, fig.alt='Configuration prompts for interactive node log the user in to the gizmo node.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.gff2211b72f_1_50")
```

## Running Interactive Commands

You can start the session by running a similar command as we used in the job we submitted via script. Echo a message by running:

```
echo "Hello, again!"
```

```{r, echo=FALSE, fig.alt='An echo message is printed on an interactive node.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.gff2211b72f_1_58")
```


## Using Pre-Installed Software Modules

Let's get a bit more advanced. We can load a preconfigured software bundle called a module. This is very convenient because it means we don't need to install anything manually! In this example, we will load a module containing [R](https://www.r-project.org/) version 4.2.0. You can learn more about what modules are available and how to request new ones for the Fred Hutch cluster [here](https://sciwiki.fredhutch.org/scicomputing/compute_scientificSoftware/).

```
ml R/4.2.0-foss-2021b
```

Next, launch R:

```
R
```

```{r, echo=FALSE, fig.alt='The R module and R sessions have been launched on the interactive node.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.gff2211b72f_1_65")
```

You can play around with R here. For example, you might run:

```
head(mtcars)
```

```{r, echo=FALSE, fig.alt='R commands on an interactive node.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.gff2211b72f_1_72")
```

Close the R session by typing:

```
q()
```

Close the interactive node by typing:

```
exit
```

```{r, echo=FALSE, fig.alt='The R session and interactive node have been exited.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.gff2211b72f_1_82")
```

<div class = "dictionary">
`grabnode` 
  
This command starts an interactive session, or node.

**CPU**

A computer component that performs and orchestrates computational tasks.

**Memory**

A computer component that stores calculations and information in the short term.
</div>

# (PART\*) Assessment {-}

# Cluster 101 Self Test
