# (PART\*) Appendix {-}

# Where to get help {#help}

The Scientific Computing group in IT manages the cluster, provides support with software, advises on data storage, and holds office hours specifically to help users. Here are some ways you can get help for your work on the cluster. 

## Send an Email {-}

The primary way you can request help for a problem is to send SciComp an email, so a ticket will be created in their tracking system.  This allows the details of the problem you're having to be sent to them so they can better help you. Submitting a good email ticket helps the SciComp Team address your needs quickly and efficiently. We suggest you submit the following information:

1.  A brief overview of what the problem is.
2.  Some specifics about the problem, such as the full text (it's ok if it's long) of any error message or terminal command, or a screen shot of the interface you were using when you had the problem. 
3.  A description of what you wanted to have happen or what your overall goal is (in case perhaps there is another strategy that might work better).

## Visit the SciWiki {-}

The SciWiki [Scientific Computing page](https://sciwiki.fredhutch.org/scicomputing/comp_index/) is full of useful tips and guides. Remember when using the search that the login "nodes" to the Fred Hutch cluster are called rhino and the cluster "nodes" are called gizmo.  

## Find Community Support on Slack {-}

Often peer to peer support can be very valuable in learning how to use and troubleshoot your work.  The [Fred Hutch Data Slack](https://fhdata.slack.com) workspace is open to all with a fredhutch.org, uw.edu, seattlechildrens.org and related institution email addresses (whi.org, scharp.org, etc). You can ask questions, find out about office hours and other live support or training events in Slack that can help you learn more about how to leverage resources at Fred Hutch to advance your science.  

# Provide Feedback {#feedback}

Please submit an issue about this course at our [GitHub repository](https://github.com/fhdsl/FH_Cluster_Guide/issues/new). You can also click the edit button on the top of the page in question.

# FAQ / Troubleshooting

## FAQ 

Here are some questions you might have.

### How can I manually install PuTTY? {#manual-putty}

<details><summary>Click to view steps</summary><p>

1. Click [here](https://www.chiark.greenend.org.uk/~sgtatham/putty/latest.html) to install the latest version of PuTTY. You will choose the 64-bit x86 installation with few exceptions.

    ```{r, echo=FALSE, fig.alt='The PuTTY latest version installation website.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g15643d101eb_4_6")
    ```

1. Click through to install via the Setup Wizard.

</p></details>

## Troubleshooting

Here are some issues you might encounter.

### `ssh: Could not resolve hostname rhino: nodename nor servname provided, or not known` {-}

<details><summary>Click to view steps</summary><p>

This error means that your computer is having trouble connecting to rhino. Ensure one of the following is true:

1. You are connected to the Fred Hutch wifi network on campus.
1. You are connected to the Fred Hutch VPN
1. You are plugged into an ethernet cable on campus that taps into the Fred Hutch network. Note that not all ethernet wall jacks have this capability, so try another jack if you are having trouble. Please email the IT helpdesk and include your office number and the number on the jack if you find a jack that isn't working.

</p></details>

### `ssh: connect to host rhino port 22: Undefined error: 0` {-}

<details><summary>Click to view steps</summary><p>

This likely indicates a disruption to your internet connection and/or VPN. Ensure you are connected to the internet and connected to the Fred Hutch network on campus or the VPN.

</p></details>

### `Connection failed` message in Cyberduck {-}

<details><summary>Click to view steps</summary><p>

This likely indicates a disruption to your internet connection and/or VPN. Ensure you are connected to the internet and connected to the Fred Hutch network on campus or the VPN.

```{r, echo=FALSE, fig.alt='Screenshot of the Cyberduck Connection failed popup windpw.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.gff2211b72f_1_16")
```

### `Invalid account or account/partition` when logging in {-}

Errors similar to this typically indicate that the account hasn’t been set up by SciComp. This is a quick fix if you [use the form mentioned in the course](01-101.html#pi-account).

</p></details>
