---
title: "FH Cluster Guide"
date: "`r format(Sys.time(), '%B %d, %Y')`"
site: bookdown::bookdown_site
documentclass: book
bibliography: book.bib
biblio-style: apalike
link-citations: yes
description: Description about Course/Book.
favicon: assets/favicon.ico
---

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), "bookdown", "knitr", "rmarkdown"
), "packages.bib")
```

# About this Course {-}

Our goal is to get you running on the Fred Hutch cluster quickly and efficiently with this quick-start guide. As a wise Drivers' Ed instructor once said, **you need to go slow to go fast**! 

In this short course, you'll invest a bit of time now to save you time and frustration down the road. Follow along at **any time on your own schedule**. We hope that the following modules will help you take advantage of the powerful resources the Fred Hutch has to offer!

We will cover how to get logged in, how to submit your first job, how to transfer files, and how to launch an interactive session.

```{r, echo=FALSE, fig.alt='Pathway of chapters covered in this course.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.gff2211b72f_1_88")
```

<!-- we would rather folks waste computing resources than not use it at all! -->
