---
title: "Fred Hutch Cluster 101"
date: "`r format(Sys.time(), '%B %d, %Y')`"
site: bookdown::bookdown_site
documentclass: book
bibliography: book.bib
biblio-style: apalike
link-citations: yes
description: Description about Course/Book.
favicon: assets/favicon.ico
---

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), "bookdown", "knitr", "rmarkdown"
), "packages.bib")
```

# About this Course {-}

Fred Hutch maintains a high performance computing cluster specifically to support research work that requires more intensive computing than your average desktop computer can handle.

Our goal for this course is to get you running on the Fred Hutch cluster **quickly and efficiently**. It is intended for everyone whether you're completely new to using a cluster, or who may have used a cluster at another institution but are new to Fred Hutch. 

The Fred Hutch cluster is supported by a group in IT called Scientific Computing, and this course was developed by the Fred Hutch Data Science Lab in collaboration with them. Please see the [author credits](#about-the-authors) for more information.

We hope that the following modules will help you take advantage of the powerful resources the Fred Hutch has to offer! This course is available in [Bookdown](https://hutchdatascience.org/FH_Cluster_Guide) and [Leanpub](https://leanpub.com/courses/fredhutch/fredhutchcluster101) formats. If you want a certificate, you need to take the Leanpub version of the course.

```{r, echo=FALSE, fig.alt='Pathway of chapters covered in this course.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.gff2211b72f_1_88")
```

# Skip to Certification {-}

Are you an experienced user? You can jump straight to the 10-question quiz by clicking the link below:

[Self-Test: Cluster 101](https://leanpub.com/courses/fredhutch/fredhutchcluster101/quizzes/self_test_101)

```{r, echo=FALSE, fig.alt='An arrow depicts jumping from the course to a Leanpub certificate.', out.width = '100%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BQxrVYdKZTbpCaF-i_q9w7s9x034lEXpQZDU-Sl09cs/edit#slide=id.g162fb43cc93_0_0")
```
